Bayesian hidden Markov model analysis of single-molecule force spectroscopy:\\ Characterizing kinetics under measurement uncertainty

### 2015-02-16 manuscripts/bayesian-hmm-method/bhmm.tex
Edits to abstract.

Single-molecule force spectroscopy has proven to be a powerful tool for studying the kinetic behavior of biomolecules.
Through application of an external force, conformational states with small or transient populations can be stabilized, allowing them to be characterized and the statistics of individual trajectories studied to provide insight into biomolecular folding and function.
Because the observed quantity (force or extension) is not necessarily an ideal reaction coordinate, individual observations cannot be uniquely associated with kinetically distinct conformations.
While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and build in detailed balance to further reduce uncertainty through physical constraints.
We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}

%\emph{Keywords: force spectroscopy; optical tweezers; single-molecule experiments; hidden Markov model (HMM); Bayesian analysis; statistical error; statistical uncertainty}

>>>>

Single-molecule force spectroscopy has proven to be a powerful tool for studying the kinetic behavior of biomolecules.
Through application of an external force, conformational states with small or transient populations can be stabilized, allowing them to be characterized and the statistics of individual trajectories studied to provide insight into biomolecular folding and function.
Because the observed quantity (force or extension) is not necessarily an ideal reaction coordinate, individual observations cannot be uniquely associated with kinetically distinct conformations.
While maximum-likelihood schemes such as hidden Markov models have solved this problem for other classes of single-molecule experiments by using temporal information to aid in the inference of a sequence of distinct conformational states, these methods do not give a clear picture of how precisely the model parameters are determined by the data due to instrument noise and finite-sample statistics, both significant problems in force spectroscopy.
{\color{blue}We solve this problem through a Bayesian extension that allows the experimental uncertainties to be directly quantified, and allow uncertainty to be further reduced through the imposition of physical priors such as the requirement that dynamics obey detailed balance.}
We illustrate the utility of this approach in characterizing the {\color{red}observed multistate kinetics of two molecules in a stationary optical trap: an RNA hairpin that exhibits three-state behavior and a protein system (apomyglobin at pH 5) that exhibits two-state behavior with highly overlapping force distributions characterizing each state.}

\emph{Keywords: force spectroscopy; optical tweezers; single-molecule experiments; hidden Markov model (HMM); Bayesian analysis; statistical error; statistical uncertainty}
